---
title: "Monthly Lab TAT Report"
author: "Your Name Here"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
 
library(tidyverse)
library(lubridate)

```

### Report date: `r format(Sys.Date(), format="%B %d %Y")`

  

```{r data, include = FALSE}
#read in data, clean up and calculate TAT in minutes
data <- read_csv("Data_Files/lab_data.csv") #read in data
glimpse(data) #take a quick look at formatting

data <- data %>%
          mutate_if(is.character, as.factor) %>%
          mutate(TAT = as.integer(difftime(RESULT_DATE, ORDERED_DATE, 
                                           units = "min")), 
                 lab.TAT = as.integer(difftime(RESULT_DATE, RECEIVED_DATE, 
                                               units = "min")),
                 month = month(RESULT_DATE, label = TRUE, abbr = TRUE),
                 year = year(RESULT_DATE)) %>%
          filter(TAT > 5, TAT < quantile(TAT, 0.99), 
                 lab.TAT > 5, lab.TAT < quantile(lab.TAT, 0.99)) #treat outliers
```


## Data  
This report describes laboratory turn-around-times (TAT) for tests ordered between `r format(min(data$ORDERED_DATE), format="%B %d %Y")` and `r format(max(data$ORDERED_DATE), format="%B %d %Y")`. TAT values less than 5 minutes or greater than the 99th percentile were excluded from calculations.  

  

## Calculations  
Turn-around-times in minutes, are calculated as follows:  
  * TAT = Result time (min) - Order time (min)  
  * Lab TAT = Result time (min) - Receive time (min)  

  

## Summary Tables for each Location by Test (All Priorities)    
```{r summary, include = FALSE}

locations <- levels(data$LOCATION_TYPE)
Summary <- vector("list", length(locations))

for (i in locations) {

  Summary[[i]] <- data %>%
    group_by(TEST, LOCATION_TYPE) %>% 
    filter(LOCATION_TYPE == i) %>%
    summarize(avg.TAT = round(mean(TAT),0),
                    avg.lab.TAT = round(mean(lab.TAT),0),
                    max.TAT = max(TAT),
                    max.labTAT = max(lab.TAT),
                    n.test = n()) %>%
    select(-LOCATION_TYPE)
}

```


```{r print tables, results='asis'}
#want to print the tables that are now items in a list (items 5-8, in this case)

for (i in 5:8) {
  
print(knitr::kable(Summary[[i]],
                   col.names = c("Test", "Avg TAT", "Avg Lab TAT",
                                    "Max TAT", "Max Lab TAT", "N"),
                   caption = names(Summary[i])))
}
```

  
  
  
## Plots of Lab TAT for Stats by Test and Location  
```{r plots}
stat.data <- data %>%
            filter(PRIORITY == "S")

ggplot(stat.data, aes(x = LOCATION_TYPE, y = lab.TAT)) +
  geom_violin() +
  geom_point(stat = "summary", fun.y = "mean", color = "red", size = 2) +
  labs(x = "", y = "Lab TAT (min)") +
  theme_classic() +
  coord_flip() + 
  facet_wrap(~TEST)
```


